---
title: "Speedtest"
author: "Wolfgang Huber"
output: 
  html_document: 
    keep_md: true
---

# Install the speedtest command line interface

The tool is called [`speedtest-cli`](https://github.com/sivel/speedtest-cli). The below shell command is for Mac OS X and homebrew. Replace by your favourite package manager.

```{sh, eval = FALSE}   
brew install speedtest-cli
```

# Set up crontab to run it in regular intervals

This works on Unix-derived systems like Linux and Mac OS X. Edit the crontab using

```{sh, eval = FALSE}
env EDITOR=nano crontab -e
```

and add a line similar to:

```{sh, eval = FALSE}
 */5 * * * * /Users/whuber/svnco/speedtest/run-speedtest-cli.sh
```

This will run the shell script `run-speedtest-cli.sh`, which is provided in a separate file, every 5 minutes. Of course you'll have to adapt the file path to your local setup. Check the documentation of `crontab` if you want to run it at different times or intervals. Make sure the file is executable, with e.g.

```{sh, eval = FALSE}
chmod 755 /Users/whuber/svnco/speedtest/run-speedtest-cli.sh
```


# Visualize the data

```{r packages, message = FALSE}
library("readr")
library("tidyr")
library("dplyr")
library("magrittr")
library("ggplot2")
```

Get the CSV header and read the CSV file
```{r read, message = FALSE}
hosts = c("spinoza", "boltzmann")
logfile = "/Users/whuber/Dropbox/speedtest/speedtest-%s.csv"
  
header = system2("speedtest-cli", args = c("--csv-header"), stdout = TRUE)  %>%
  strsplit(split = ",") %>% `[[`(1)

st = lapply(hosts, function(h) {
  read_csv(sprintf(logfile, h), col_names = FALSE) %>%
  `colnames<-`(header)  %>% 
  mutate(hostname = h)
}) %>% bind_rows 

stopifnot(all(is.finite(st$Timestamp)))
is_fin = is.finite(st$Download)
stopifnot(identical(is_fin, is.finite(st$Upload)),
          identical(is_fin, is.finite(st$Ping)))
with(st,
  table(ifelse(is.na(Download), "NA", ifelse(Download==0, "0", ">0")), 
        hostname))

variables = c("Download (MB/s)", "Upload (MB/s)", "log10 Ping (ms)")
st %<>% mutate(
  `Download (MB/s)` = Download / 2^23,
  `Upload (MB/s)` = Upload / 2^23,
  `log10 Ping (ms)` = log10(Ping)) %>% 
   pivot_longer(cols = all_of(variables))
st$name %<>% factor(levels = variables)
```

Stratify by  "IP Address", which indicates whether the measurement was taken in Unitymedia broadband, LTE Telekom or at work at EMBL.

The `stopifnot` call makes sure that all instances of `IP Address` that are not `NA` are one of the three addresses below. 
Subsequently we assume that the cases where `IP Address` is `NA` were at home, i.e. in the Unitymedia network.

```{r servers}
servers = c(Unitymedia = "176.199.210.217", EMBL = "194.94.44.220", Telekom = "80.187.97.51")
mt = match(st$`IP Address`, servers)
stopifnot(all( (!is.na(mt)) | is.na(st$`IP Address`) ))
st$network = names(servers)[ifelse(is.na(mt), 1, mt)]
```

```{r speedtestplot, fig.dim = c(9, 8.1), results = "hide"}
makeplot = function(x, thenetwork, from = "2020-01-01", to = "2020-12-31") {
  x = filter(x, Timestamp >= from & Timestamp <= to & network == thenetwork) 
  print(ggplot(x, aes(x = Timestamp, y = ifelse(is.na(value), 0, value))) + 
    # Sadly, it seems that the below timezone setting is ignored by ggplot2
    scale_x_datetime(timezone = "CEST") + 
    xlab("Time") + ylab("Value") + 
    geom_point(aes(col = hostname, shape = is.na(value)), size = 0.5) + 
    facet_grid(rows = vars(name), scales = "free_y") +
    scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 4)) +
    theme(legend.position="bottom") + scale_colour_brewer(palette = "Set1") +
    guides(`is.na(value)` = FALSE)) 
  dev.copy(pdf, file = sprintf("speedtest_%s.pdf", thenetwork), width = 16, height = 8)
}
makeplot(st, thenetwork = "Unitymedia")
```